A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-02-28 analysis/paper/paper.Rmd
Adding references

The post-glacial relative sea-level fall that characterises large areas of Fennoscandia is fundamental to its archaeology. This follows not only from the dramatic changes to the landscape that this process has represented throughout prehistory, but also from the fact that if archaeological phenomena were situated close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level. This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, various harbour and sea-side constructions and, as is the focus of this study, Stone Age sites [e.g. @akerlund1996; @bjerck2005; @nordqvist1995; @sognnes2003; @tallavaara2020].

@@

The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways. The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87; but see @gundersen2013; @mjærum2018; @schulke2020]. This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020]. The introduction of a comprehensive Neolithic cultural package, including a shift to agro-pastoralism and the introduction of the farm, led the site locations to be more withdrawn from the shoreline [e.g. @bakka1971; @prescott2012; @ostmo2008, 223]. That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods [@glorstad2011a]. At an earlier stage, at the transition to the Early Neolithic (c. 3900 BCE), pottery is introduced to the sites, and there are some indications of an initial uptake of agriculture at some sites in the Oslo fjord region. However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955]. @nielsen2021b has recently argued that the initial uptake of agriculture in the Early Neolithic is combined with a more complex settlement pattern, and that a simple forager/farming dichotomy would underplay the variation present in the Early and Middle Neolithic data [see also e.g. @amundsen2006; @solheim2012, 74]. Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line from around 3900 BCE, and a decisive shift around 2400 BCE.

Based on the generally accepted premise that most pre-Late Neolithic sites in south-eastern Norway located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise. This is for example common in survey projects, which are often guided by both a digital and mental reconstruction of past sea-levels [e.g. @berg-hansen2009; ] . Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves it is common to assume a shore-bound site location, even if the typologically informed date-span is too wide to decisively verify this. It is also common to combine this with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was not shore bound [e.g. @jaksland2014; @nummedal1923]. This can for example pertain to accessibility. If the site is situated on a ledge in a steep and jagged area of the present day landscape, or if it would have been located on an island when the sea-level was higher, it would make intuitive sense that the site was in use when the ocean reached closer to its elevation, as the site would have been accessible by means of watercraft. Although it appears that the arguments for such site locations are sensible and can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are few, and are typically conducted on a site-by-site or project-by-project basis, meaning the basis for assuming shore-bound site location is largely a mosaic of smaller scale investigations.

@@

The employed shoreline displacement curves are all based on the so-called isolation basin method [@romundset2011]. This involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments in the cores. Each curve is thus construed from a series of cored basins located at different elevations, each representing a high precision sea-level index point. Furthermore, to minimise the impact of variable uplift rates, the basins are located in a as constrained area of the landscape as possible. Following from the morphology of the retreating ice sheet, the uplift is more severe towards the north-east, meaning that this needs to be adjusted for in the case that any basins are located any significant distance from a common isobase perpendicular to this gradient. The confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters. The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, to name but a few [e.g. @romundset2019; @romundset2011]. For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications.

@@

This process was repeated 1000 times for each phase for each site. The choice of 1000 simulation runs follows from an evaluation of when the mean distances between site and shoreline converged when running 5000 iterations of the simulation on the site Hovland 5, available in the supplementary material. Hovland 5 was chosen for this evaluation as it has a fairly uncertain date range, and is located in area of quite complex topography. At 1000 simulation runs the analysis of each site took up to a few hours, depending on the distance to the simulated shorelines and the necessary size of the window of analysis.

@@

An exponential function has been fit to the distributions for vertical distance using maximum likelihood. While it makes theoretical sense thata process of exponential decay explains this relationship, it  is clear that this does not perfectly match the data. However, this can in part be related to methodological factors, where the accumulation of distances-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero. Having derived a exponential decay function for describing the vertical distance between sites and shoreline, this can be combined with the displacement data to provide a method for shoreline dating that takes this into account.

@@

```{r shoredate, echo = FALSE, fig.cap = "The result of backwards shoreline dating the sites with radiocarbon dates corresponding to the artefact inventory. The sites are dated using the mean elevation of the site polygon and by employing the exponential decay function identified in Figure 7 for characterising the distance between sites and shoreline. A) The shoreline dates in gray are plotted against the oldest radiocarbon-informed phase for each site in black. The red line indicates the 95\\% probability range for the shoreline date. B) Using the method of Parnell et al. (2008)", out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/shoredate.png"))
```

@@

In Figure \@ref(fig:redate)), excavated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites are also subjected to the method for shoreline dating outlined above. These are compared to the dates originally proposed in the excavation reports for the sites. To avoid issues with the DTM, the sites are dates based on the mean elevation of those provided for each site in the reports. This is an illustrative, but unfair exercise for a few reasons. First of all the dates provided in the reports are typically stated to be a very rough estimate, and are sometimes given as a point estimate with an implied or explicit, but undefined uncertainty range. Secondly, seeing as these reports are from various dates in time, many are based on now outdated data on RSL-change. Finally, they are often only meant to indicate a lower bound for when the sites could have been in use. I do, however, think that the results indicate that shoreline dating, in general, has been applied with a fairly reasonable degree of success, seeing as these dates have typically been interpreted and informed research in an approximate manner. The results from applying the method devised here again highlights the spatial and temporal contingency of the method. In some cases the method provides a very precise date range and in others it offers little more than a *terminus post quem*.  Furthermore, a few of the date ranges extend well beyond the Stone Age and could be severely and securely constrained with only cursory reference to typology. This thus points to another way that the dating method could be improved. In the above example from Brunlanes it was shown how this could conceivably be done by evaluating the geographical location of the sites. Another possibility is of course to draw on other temporal data, such as typological indicators or radiocarbon dates with which a shoreline date could be combined, for example within a Bayesian framework.

@@

The most immediate contribution of this paper is what must be considered a confirmation of previous research into the relation between coastal Norwegian Stone Age sites and the prehistoric shoreline. This is indicated by the close relationship between sites and the shoreline up until the transition to the Neolithic at c. 3900 BCE after which a couple of sites become situated some distance from the shoreline, and the more decisive break at the transition to the Late Neolithic at c. 2400 BCE -- in clear agreement with the literature. Based on these findings, a refined method for the shoreline dating of Stone Age sites has been proposed. This involves both taking the distance between sites and the isobases of the displacement curves into consideration when dating the sites, and implementing formula x to account for the distance between the sites and the shoreline. When no other information is available, it can at present be recommended to use the empirically derived exponential decay ratios identified in Figure to characterise this relationship. However, the accuracy of the method can be improved by including more information, both with reference to the topographic location of the site and other temporal data such as typological indicators in the artefact inventories. The precision of the method is, as shown above, both geographically and temporally contingent due to the trajectory of RSL-change, where older sites situated towards the north-east in the study area will get a more accurate date than younger sites located towards the south-west. The impact of such additional information will therefore also vary.

Future investigations and radiocarbon dates from Stone Age sites in the region can not only be used to further test and adjust the findings reported here, but could also lay the foundations for refining the method. Given it's behavioural nature, it would for example seem likely that dimensions such as the nature and purpose of visits to the sites will have implications for how close to the shoreline they were located. Identifying such subsets of sites will thus potentially further refine the method, allowing it to be adjusted depending on the site being dated. Furthermore, other dimensions related to the topographic location of the sites could be similarly explored. This for example pertains to the exposure of sites to wave action, which is likely to have been of concern [@roalkvam2020], and which presumably has implications for how close to the shoreline people settled [@blankholm2020]. If such locational patterns can be discerned, this will also be useful for improving the shoreline dating of sites which have only been surveyed and where little information on the nature of the occupation of the site is available.

It is also worth highlithing some sources of likely variation and uncertainty have not been considered here. First of all the DTM has only been corrected for major modern disturbances. This means that erosion, although likely not that prevalent, has not been taken into account. Secondly, the DTM has a vertical error [cf. @lewis2021].

@@

@@

This quantification was achieved by means of Monte Carlo-simulation, integrating multiple sources of spatio-temporal uncertainty. Here the approach was simply used to inform the question of the distance between sites and shorelines. However, this method can be extended to a wide range of use-cases where one needs to visualise and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.

>>>>

The post-glacial relative sea-level fall that characterises large areas of Fennoscandia is fundamental to its archaeology. This follows not only from the dramatic changes to the landscape that this process has represented throughout prehistory, but also from the fact that if archaeological phenomena were situated close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level. This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, various harbour and sea-side constructions and, as is the focus of this study, Stone Age sites [e.g. @akerlund1996; @bjerck2005; @nordqvist1995; @sognnes2003; @tallavaara2020; @schmitt2009; @rosentau2013].

@@

The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways. The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87; see also discussion in @gundersen2013; @mjærum2018; @schulke2020]. This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020]. The introduction of a comprehensive Neolithic cultural package, including a shift to agro-pastoralism and the introduction of the farm, led the site locations to be more withdrawn from the shoreline [e.g. @bakka1971; @prescott2012; @ostmo2008, 223]. That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods [@glorstad2011a]. At an earlier stage, at the transition to the Early Neolithic (c. 3900 BCE), pottery is introduced to the sites, and there are some indications of an initial uptake of agriculture at some sites in the Oslo fjord region. However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955; @ostmo1988, 225--227]. @nielsen2021b has recently argued that the initial uptake of agriculture in the Early Neolithic is combined with a more complex settlement pattern, and that a simple forager/farming dichotomy would underplay the variation present in the Early and Middle Neolithic data [see also e.g. @amundsen2006; @solheim2012, 74; @ostmo1988]. Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line from around 3900 BCE, and a decisive shift around 2400 BCE.

Based on the generally accepted premise that most pre-Late Neolithic sites in south-eastern Norway located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise. This is for example common in survey projects, which are often guided by both a digital and mental reconstruction of past sea-levels [e.g. @berg-hansen2009; @eskeland2017]. Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves it is common to assume a shore-bound site location, even if the typologically informed date-span is too wide to decisively verify this. It is also common to combine this with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was not shore bound [e.g. @jaksland2014; @nummedal1923]. This can for example pertain to accessibility. If the site is situated on a ledge in a steep and jagged area of the present day landscape it would make intuitive sense that the site was in use when the ocean reached closer to its elevation, as the site would have been accessible by means of watercraft. Although it appears that the arguments for such site locations are sensible and can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are few [see also @ilves2011], and are typically conducted on a site-by-site or project-by-project basis, meaning the assumed shore-bound site location is largely based on a mosaic of smaller scale investigations.

@@

The employed shoreline displacement curves are all based on the so-called isolation basin method [e.g. @kjemperud1986; @romundset2011]. This involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments in the cores. Each curve is thus construed from a series of cored basins located at different elevations, each representing a high precision sea-level index point (SLIP). Furthermore, to minimise the impact of variable uplift rates, the basins are located in a as constrained area of the landscape as possible. Following from the morphology of the retreating ice sheet, the uplift is more severe towards the north-east, meaning that this needs to be adjusted for in the case that any basins are located any significant distance from a common isobase perpendicular to this gradient. The SLIPs from the isolaton basin method indicate the highest astronomical tide at the time of isolation. This is adjusted to mean sea-level based on the present day tidal range, which is assumed to have been the same throughout the Holocene [@sorensen2014, 44]. Furthermore, the confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters. The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, to name but a few [e.g. @romundset2019; @romundset2011]. For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications.

@@

This process was repeated 1000 times for each phase for each site. The choice of 1000 simulation runs follows from an evaluation of when the mean distances between site and shoreline converged when running 5000 iterations of the simulation on the site Hovland 5, available in the supplementary material [cf. @crema2010, 1125]. Hovland 5 was chosen for this evaluation as it has a fairly uncertain date range, and is located in area of quite complex topography. At 1000 simulation runs the analysis of each site took up to a few hours, depending on the distance to the simulated shorelines and the necessary size of the window of analysis.

@@

An exponential function has been fit to the distributions for vertical distance using maximum likelihood. While it makes theoretical sense that a process of exponential decay explains this relationship, it  is clear that this does not perfectly match the data. However, this can in part be related to methodological factors, where the accumulation of distances-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero. Having derived an exponential decay function for describing the vertical distance between sites and shoreline, this can be combined with the displacement data to provide a method for shoreline dating that takes this into account:

@@

```{r shoredate, echo = FALSE, fig.cap = "The result of backwards shoreline dating the sites with radiocarbon dates corresponding to the artefact inventory. The sites are dated using the mean elevation of the site polygon and by employing the exponential decay function identified in Figure 7 for characterising the distance between sites and shoreline. A) The shoreline dates in gray are plotted against the oldest radiocarbon-informed phase for each site in black. The red line indicates the 95\\% probability range for the shoreline date. B) Using the method of Parnell et al. (2008), the synchroneity of the dates is evaluated ", out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/shoredate.png"))
```

@@

In Figure \@ref(fig:redate), excavated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites are also subjected to the method for shoreline dating outlined above. These are compared to the dates originally proposed in the excavation reports for the sites. To avoid issues with the DTM, the sites are dates based on the mean elevation of those provided for each site in the reports. This is an illustrative, but unfair exercise for a few reasons. First of all the dates provided in the reports are typically stated to be a very rough estimate, and are sometimes given as a point estimate with an implied or explicit, but undefined uncertainty range. Secondly, seeing as these reports are from various dates in time, many are based on now outdated data on RSL-change. Finally, they are often only meant to indicate a lower bound for when the sites could have been in use. I do, however, think that the results indicate that shoreline dating, in general, has been applied with a fairly reasonable degree of success, seeing as these dates have typically been interpreted and informed research in an approximate manner. The results from applying the method devised here again highlights the spatial and temporal contingency of the method. In some cases the method provides a very precise date range and in others it offers little more than a *terminus post quem*.  Furthermore, a few of the date ranges extend well beyond the Stone Age and could be severely and securely constrained with only cursory reference to typology. This thus points to another way that the dating method could be improved. In the above example from Brunlanes it was shown how this could conceivably be done by evaluating the geographical location of the sites. Another possibility is of course to draw on other temporal data, such as typological indicators or radiocarbon dates with which a shoreline date could be combined, also within a Bayesian framework.

@@

It is worth mentioning some sources of likely variation and uncertainty that have not been considered here. First of all the DTM has only been corrected for major modern disturbances. This means that erosion, although likely not that prevalent, has not been taken into account. Secondly, the DTM has a vertical error [cf. @lewis2021].

@@

The most immediate contribution of this paper is what must be considered a confirmation of previous research into the relation between coastal Norwegian Stone Age sites and the prehistoric shoreline. This is indicated by the close relationship between sites and the shoreline up until the transition to the Neolithic at c. 3900 BCE, after which a couple of sites become situated some distance from the shoreline, and the more decisive break at the transition to the Late Neolithic at c. 2400 BCE -- in clear agreement with the literature. Furthermore, based on the quantative nature of these findings, a refined method for the shoreline dating of Stone Age sites has been proposed. This involves both taking the distance between sites and the isobases of the displacement curves into consideration when dating the sites, and implementing formula x to account for the distance between the sites and the shoreline. When no other information is available, it can at present be recommended to use the empirically derived exponential decay ratios identified in Figure \@ref(fig:redate) to characterise this relationship. However, the accuracy of the method can be improved by including more information, both with reference to the topographic location of the sites and other temporal data such as typological indicators in the artefact inventories. The precision of the method is, as shown above, both geographically and temporally contingent due to the trajectory of RSL-change, where older sites situated towards the north-east in the study area will get a more accurate date than younger sites located towards the south-west. The impact of such additional information will therefore also vary.

Future investigations and radiocarbon dates from Stone Age sites in the region can not only be used to further test and adjust the findings reported here, but a larger sample size could also lay the foundations for refining the method by identifying subsets of sites for which the application of the method could be adjusted. Given it's behavioural nature, it would for example seem likely that dimensions such as the nature and purpose of visits to the sites will have implications for how close to the shoreline they were located. Furthermore, other dimensions related to the topographic location of the sites could be similarly explored. This for example pertains to the exposure of sites to wave action, which is likely to have been of concern [e.g. @roalkvam2020], and which presumably has implications for how close to the shoreline people settled [@blankholm2020]. If such locational patterns can be discerned, this will also be useful for improving the shoreline dating of sites which have only been surveyed and where little information on the nature of the occupation of the site is available.

Finally, the analysis done here was Monte Carlo-simulation, integrating multiple sources of spatio-temporal uncertainty. Here the approach was simply used to inform the question of the distance between sites and shorelines. However, this method can be extended to a wide range of use-cases where one needs to visualise and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.

@@
